% Backmatter sections for the manuscript

\section*{Acknowledgements}
SS was supported by an R+ Postdoctoral Fellowship. 

\section*{Code and Data Availability}
The code for the AIME framework is available in a public GitHub repository at \url{https://github.com/s-spillias/EMs-with-LLMs}. The repository includes all scripts necessary to reproduce the results presented in this paper, including the genetic algorithm implementation, model evaluation tools, and analysis scripts. The time series data used in the case studies are also available in the repository. The software is released under an MIT license.

\section*{Declaration on Generative AI Usage}
During the preparation of this manuscript, we used Claude-3.5-Sonnet, a large language model, to assist with code documentation, manuscript formatting, and language editing. The scientific content, analyses, interpretations, and conclusions presented in this paper were developed and validated by the human authors. 

\section*{Author Contributions}
SS: Conceptualization, Methodology, Software, Data curation, Formal analysis, Writing - original draft, Writing - review \& editing
JR: Software, Methodology, Data curation, Writing - review \& editing
FB: Formal analysis, Supervision, Writing - review \& editing
BF: Formal analysis, Supervision, Writing - review \& editing
RT: Supervision, Writing - review \& editing
MG: Software, Methodology, Writing - review \& editing
SY: Software, Methodology, Writing - review \& editing

\section*{Competing Interests}
The authors declare no competing interests.

\section*{Data Availability}
The datasets generated and analyzed during the current study are available in the GitHub repository \url{https://github.com/s-spillias/EMs-with-LLMs}. Additional data that support the findings of this study are available from the corresponding author upon reasonable request.

