%========================================================
% Readable Equations + Parameter Tables for Best Models
%========================================================

\section{Best Performing Models for CoTS Case Study}
\label{sec:best_models_cots}

\subsection{openrouter:openai/gpt-5 Model (CoTS)}
\paragraph{States and data.}
Adults \(C_t\) (\si{ind.m^{-2}}), juveniles \(J_t\) (\si{ind.m^{-2}}), fast coral \(F_t\) (\%), slow coral \(S_t\) (\%), sea-surface temperature \(\mathrm{SST}_t\) (\si{\degreeC}), immigration \(I^{\text{cots}}_t\) (\si{ind.m^{-2}.yr^{-1}}).

\paragraph{Process model (discrete yearly time, using \(t-1\) drivers).}
\begin{align}
f_{\text{Allee}}(C) &= \frac{1}{1 + \exp\!\big[-k_{\text{allee}}(C - C_{\text{allee}})\big]} \\
f_{T,\text{rec}}(\mathrm{SST}) &= \exp\!\big[-\beta_{\text{rec}}(\mathrm{SST} - T_{\text{opt,rec}})^2\big] \\
\text{stock}(C) &= \frac{C^{\phi}}{1 + C/C_{\text{sat,rec}}} \\
f_{\text{food}}(\mathrm{food}) &= 
\begin{cases}
\displaystyle \frac{\mathrm{food}}{K_{\text{food}} + \mathrm{food}}, & \text{if driver available}\\[6pt]
1, & \text{otherwise}
\end{cases} \\[4pt]
\mathrm{Rec}_t &= \alpha_{\text{rec}}\cdot \text{stock}(C_{t-1}) \cdot f_{\text{Allee}}(C_{t-1})\cdot f_{T,\text{rec}}(\mathrm{SST}_{t-1})\cdot f_{\text{food}}(\mathrm{food}_{t-1}) + I^{\text{cots}}_{t-1} \\[4pt]
\mathrm{Mort}^{\text{adult}}_{t} &= (\mu_C + \gamma_C\, C_{t-1})\, C_{t-1} \\
\mathrm{Mat}_t &= m_J\, J_{t-1}, \qquad \mathrm{Mort}^{\text{juv}}_t = \mu_J\, J_{t-1} \\[4pt]
C_t &= \max\!\big\{0,\; C_{t-1} + \mathrm{Mat}_t - \mathrm{Mort}^{\text{adult}}_{t}\big\} \\
J_t &= \max\!\big\{0,\; J_{t-1} + \mathrm{Rec}_t - \mathrm{Mat}_t - \mathrm{Mort}^{\text{juv}}_{t}\big\} \\[4pt]
\text{heat}_t &= \max\!\big\{0,\; \mathrm{SST}_{t-1} - T_{\text{opt,bleach}}\big\}, \quad
\text{growth}_{\text{mod},t} = \exp(-\beta_{\text{bleach}}\cdot \text{heat}_t) \\[4pt]
\text{space}_t &= 1 - \frac{F_{t-1}+S_{t-1}}{K_{\text{tot}}} \\[4pt]
G^{\text{fast}}_t &= r_F\, F_{t-1}\cdot \text{space}_t \cdot \text{growth}_{\text{mod},t}, \qquad
G^{\text{slow}}_t = r_S\, S_{t-1}\cdot \text{space}_t \cdot \text{growth}_{\text{mod},t} \\[4pt]
B^{\text{fast}}_t &= m_{\text{bleach}F}\cdot \text{heat}_t \cdot F_{t-1}, \qquad
B^{\text{slow}}_t = m_{\text{bleach}S}\cdot \text{heat}_t \cdot S_{t-1} \\[4pt]
\text{denom}_t &= 1 + h\big(a_F F_{t-1}^{\eta_F} + a_S S_{t-1}^{\eta_S}\big) \\
\mathrm{Cons}^{\text{fast}}_t &= q_F \frac{a_F F_{t-1}^{\eta_F} C_{t-1}}{\text{denom}_t}, \qquad
\mathrm{Cons}^{\text{slow}}_t = q_S \frac{a_S S_{t-1}^{\eta_S} C_{t-1}}{\text{denom}_t} \\[4pt]
F_t &= \mathrm{clamp}_{[0,100]}\!\big(F_{t-1} + G^{\text{fast}}_t - \mathrm{Cons}^{\text{fast}}_t - B^{\text{fast}}_t\big) \\
S_t &= \mathrm{clamp}_{[0,100]}\!\big(S_{t-1} + G^{\text{slow}}_t - \mathrm{Cons}^{\text{slow}}_t - B^{\text{slow}}_t\big)
\end{align}

\paragraph{Observation model.}
\begin{align}
\log C^{\text{obs}}_t &\sim \mathcal{N}\!\big(\log C_t,\; \sigma_{\text{cots}}^2\big), \quad \text{with Jacobian } +\log C^{\text{obs}}_t \\
\mathrm{logit}\big(F^{\text{obs}}_t\big) &\sim \mathcal{N}\!\big(\mathrm{logit}(F_t),\; \sigma_{\text{fast}}^2\big) \\
\mathrm{logit}\big(S^{\text{obs}}_t\big) &\sim \mathcal{N}\!\big(\mathrm{logit}(S_t),\; \sigma_{\text{slow}}^2\big)
\end{align}

\paragraph{Parameters.}
\begin{longtable}{@{}l l l X l l@{}}
\toprule
\textbf{Symbol} & \textbf{Name} & \textbf{Units} & \textbf{Description} & \textbf{Value} & \textbf{Bounds / Source}\\
\midrule
\endhead
$C_0$ & Initial adult COTS & \si{ind.m^{-2}} & Initial condition at \(t=0\) & 0.1 & [0, 50] (init est.)\\
$J_0$ & Initial juveniles & \si{ind.m^{-2}} & Initial juvenile pool & 0.1 & [0, 50] (init est.)\\
$F_0$ & Initial fast coral & \% & Initial fast coral cover & 30.0 & [0, 100] (init est.)\\
$S_0$ & Initial slow coral & \% & Initial slow coral cover & 30.0 & [0, 100] (init est.)\\
$\alpha_{\text{rec}}$ & Recruitment scaling & \si{ind.m^{-2}.yr^{-1}} & Productivity scaling to juveniles & 1.0 & [0, 10] (init est.)\\
$\phi$ & Fecundity exponent & -- & Density exponent in fecundity & 1.5 & [1, 3] (init est.)\\
$k_{\text{allee}}$ & Allee steepness & \si{m^{2}.ind^{-1}} & Logistic steepness of Allee effect & 2.0 & [0.01, 20] (init est.)\\
$C_{\text{allee}}$ & Allee threshold & \si{ind.m^{-2}} & Density threshold for mating success & 0.2 & [0, 5] (init est.)\\
$C_{\text{sat,rec}}$ & Stock–recruit taper & \si{ind.m^{-2}} & Beverton–Holt scale (taper) & 2.0 & [0.01, 50] (improvement)\\
$\mu_C$ & Adult baseline mort. & \si{yr^{-1}} & Baseline adult mortality & 0.6 & [0, 3] (init est.)\\
$\gamma_C$ & Density mort. coeff. & \si{m^{2}.ind^{-1}.yr^{-1}} & Density-dependent mortality & 0.5 & [0, 10] (init est.)\\
$m_J$ & Maturation fraction & \si{yr^{-1}} & Annual juvenile maturation & 0.5 & [0, 1] (init est.)\\
$\mu_J$ & Juvenile mort. & \si{yr^{-1}} & Juvenile mortality & 0.5 & [0, 1] (init est.)\\
$T_{\text{opt,rec}}$ & Opt. SST (recruit) & \si{\degreeC} & Optimal SST for recruitment & 26.5 & [20, 34] (lit.)\\
$\beta_{\text{rec}}$ & Temp curvature & \si{\degreeC^{-2}} & Gaussian curvature of temp effect & 0.2 & [0, 2] (init est.)\\
$K_{\text{food}}$ & Food half-sat. & (driver units) & Monod half-saturation (if activated) & 1.0 & [0.001, 100] (scaffold)\\
$T_{\text{opt,bleach}}$ & Bleach onset SST & \si{\degreeC} & Threshold for bleaching stress & 32.65 & [31.0, 34.3] (lit.)\\
$\beta_{\text{bleach}}$ & Growth reduction & -- & Growth reduction under heat & 0.5 & [0, 5] (init est.)\\
$m_{\text{bleach}F}$ & Fast bleaching loss & \si{yr^{-1}.\degreeC^{-1}} & Proportional loss per \si{\degreeC} & 0.2 & [0, 2] (init est.)\\
$m_{\text{bleach}S}$ & Slow bleaching loss & \si{yr^{-1}.\degreeC^{-1}} & Proportional loss per \si{\degreeC} & 0.1 & [0, 2] (init est.)\\
$r_F$ & Fast coral regrowth & \si{yr^{-1}} & Intrinsic regrowth (fast) & 0.5 & [0, 2] (lit./init)\\
$r_S$ & Slow coral regrowth & \si{yr^{-1}} & Intrinsic regrowth (slow) & 0.2 & [0, 2] (lit./init)\\
$K_{\text{tot}}$ & Total coral capacity & \% & Carrying capacity (F+S) & 70.0 & [10, 100] (lit./init)\\
$a_F,a_S$ & Attack params & see code & Encounter/attack on corals & 0.02; 0.01 & [0, 1] (init est.)\\
$\eta_F,\eta_S$ & Shape exponents & -- & Type-III blend exponents & 1.5; 1.2 & [1, 3] (init est.)\\
$h$ & Handling scaler & \si{yr.\%^{-1}} & Satiation scaler & 0.02 & [0, 1] (init est.)\\
$q_F,q_S$ & Conversion efficiencies & -- (0–1) & % cover loss per feeding & 0.8; 0.5 & [0, 1] (lit./init)\\
$\sigma_{\text{cots}}$ & COTS log-SD & log-space SD & Observation SD (COTS) & 0.5 & [0.01, 2] (init est.)\\
$\sigma_{\text{fast}}$ & Fast logit-SD & logit SD & Observation SD (fast coral) & 0.3 & [0.01, 2] (init est.)\\
$\sigma_{\text{slow}}$ & Slow logit-SD & logit SD & Observation SD (slow coral) & 0.3 & [0.01, 2] (init est.)\\
\bottomrule
\end{longtable}

\medskip
\noindent\emph{Notes: All equations and parameter values above are taken directly from your file’s model and parameter listings.} 

\subsection{openrouter:anthropic/claude-sonnet-4.5 Model (CoTS)}
\paragraph{Key features.}
Type-II predation on corals, Allee effects, immigration, logistic adult growth, and \textbf{unimodal substrate–food favorability} for recruitment peaking at intermediate total coral cover.

\paragraph{Process model (selected equations).}
\begin{align}
\text{Total coral: } T_{t-1} &= F_{t-1} + S_{t-1} + \varepsilon \\
\text{Temp effect: } f_T(\mathrm{SST}_t) &= \exp\!\left[-\tfrac12\left(\frac{\mathrm{SST}_t - T_{\text{opt}}}{\text{width}}\right)^2\right] \\[4pt]
\text{Substrate–food favorability: } f_{\text{SF}}(T) &= \frac{T}{T_{\text{opt,coral}}}\,\exp\!\left(1 - \frac{T}{T_{\text{opt,coral}}}\right), \quad \text{bounded to }[0,1] \\[4pt]
\text{Composite favorability: } FAV_t &= f_T(\mathrm{SST}_t)\cdot \frac{I^{\text{cots}}_t}{\max I^{\text{cots}}}\cdot f_{\text{SF}}(T_{t-1}) \\[4pt]
\text{Recruit activation: } A_t &= \frac{1}{1+\exp\big[-20\,(FAV_t - \tau)\big]} \\[4pt]
\text{Recruit pulse: } R^{\text{pulse}}_{t} &= R_{\max}\cdot A_t\cdot FAV_t \\[4pt]
\text{Adult growth: } \Delta C_t &= r_C\,C_{t-1}\cdot f_{\text{Allee}}(C_{t-1})\cdot f_T(\mathrm{SST}_t)\cdot (1 - C_{t-1}/K_C)
+ \eta_{\text{conv}}\cdot \text{(total consumption)} \\
\text{Mortality: } M_t &= (\mu_0 + \mu_{\text{dd}} C_{t-1})\cdot \big[1 + 2\exp(-T_{t-1}/5)\big]
\end{align}
\(\text{Type-II consumption on fast/slow corals and coral logistic dynamics with temperature stress are as in the implementation.}\)

\paragraph{Observation model.}
\begin{align}
\log C^{\text{obs}}_t &\sim \mathcal{N}\!\big(\log C_t,\; \sigma_{\text{cots}}^2\big),\qquad
F^{\text{obs}}_t \sim \mathcal{N}(F_t,\; \sigma_{\text{fast}}^2),\qquad
S^{\text{obs}}_t \sim \mathcal{N}(S_t,\; \sigma_{\text{slow}}^2).
\end{align}

\paragraph{Parameters (selected; log-parameters shown on natural meaning).}
\begin{longtable}{@{}l l l X l l@{}}
\toprule
\textbf{Symbol} & \textbf{Name} & \textbf{Units} & \textbf{Description} & \textbf{Value} & \textbf{Bounds / Source}\\
\midrule
\endhead
$r_C$ & Intrinsic COTS growth & \si{yr^{-1}} & Adults’ logistic growth (unlogged) & \(\exp(\text{log\_r\_cots}) \approx 0.5\) & log in file (lit.)\\
$K_C$ & COTS carrying capacity & \si{ind.m^{-2}} & Adults’ carrying capacity & \(\exp(\text{log\_K\_cots}) \approx 5\) & log in file (lit.)\\
$C_{\text{allee}}$ & Allee threshold & \si{ind.m^{-2}} & Half-saturation for Allee & \(\exp(\text{log\_allee\_threshold}) \approx 0.2\) & log in file (lit.)\\
$\mu_0$ & Baseline mortality & \si{yr^{-1}} & Baseline adult mortality & \(\exp(\text{log\_mort\_base}) \approx 0.3\) & log bounds (lit.)\\
$\mu_{\text{dd}}$ & Density mortality coeff. & \si{m^{2}.ind^{-1}.yr^{-1}} & Density-dependent mortality & \(\exp(\text{log\_mort\_density}) \approx 0.2\) & log bounds (lit.)\\
$T_{\text{opt}}$ & Opt. recruitment SST & \si{\degreeC} & Optimal temperature & \(\exp(\text{log\_temp\_opt}) \approx 28\) & log bounds (lit.)\\
$\text{width}$ & Temp tolerance & \si{\degreeC} & Gaussian width & \(\exp(\text{log\_temp\_width}) \approx 2\) & (lit.)\\
$R_{\max}$ & Max pulse recruitment & \si{ind.m^{-2}.yr^{-1}} & Peak pulse flux & \(\exp(\text{log\_recruit\_max}) \approx 1\) & (init est.)\\
$\tau$ & Pulse threshold & -- & Favorability threshold & 0.6 & [0.4, 0.8] (init est.)\\
$T_{\text{opt,coral}}$ & Opt. total coral & \% & Peak of unimodal substrate–food & \(\exp(\text{log\_optimal\_coral}) \approx 30\) & [12–50\%] (lit.)\\
\multicolumn{6}{l}{\emph{Predation, handling and observation SD parameters as listed in the file (all log-transformed where applicable).}}\\
\bottomrule
\end{longtable}

\subsection{openrouter:google/gemini-2.5-pro Model (CoTS)}
\paragraph{Process and observation (compact).}
\begin{align}
\text{Predation (Type II):}\quad
\mathrm{Cons}^{\text{fast}}_t &= \frac{a_F\,F_{t-1}\,C_{t-1}}{1 + a_F h F_{t-1} + \varepsilon},\quad
\mathrm{Cons}^{\text{slow}}_t = \frac{a_S\,S_{t-1}\,C_{t-1}}{1 + a_S h S_{t-1} + \varepsilon} \\[4pt]
\text{Bleaching mortality:}\quad
\mathrm{Bleach}_X &= \frac{m_X}{1 + \exp[-k_{\text{bleach}}(\mathrm{SST}_t - T_{\text{bleach},X})]}\cdot X_{t-1},\quad X\in\{F,S\} \\[4pt]
\text{Corals:}\quad F_t &= F_{t-1} + r_F F_{t-1}\Big(1 - \frac{F_{t-1}+S_{t-1}}{K_{\text{coral}}}\Big) - \mathrm{Cons}^{\text{fast}}_t - \mathrm{Bleach}_F \\
S_t &= S_{t-1} + r_S S_{t-1}\Big(1 - \frac{F_{t-1}+S_{t-1}}{K_{\text{coral}}}\Big) - \mathrm{Cons}^{\text{slow}}_t - \mathrm{Bleach}_S \\[4pt]
\text{Temperature-modulated reproduction:}\quad
f_T(\mathrm{SST}_t) &= \exp\!\Big[-\frac{(\mathrm{SST}_t - T_{\text{opt,cots}})^2}{2T_{\sigma,\text{cots}}^2}\Big] \\[4pt]
\text{COTS:}\quad
C_t &= C_{t-1} + \Big[e_F\,\mathrm{Cons}^{\text{fast}}_t + e_S\,\mathrm{Cons}^{\text{slow}}_t\Big]\cdot \frac{C_{t-1}}{C_{t-1}+A_{\text{allee}}}\cdot f_T(\mathrm{SST}_t) \\
&\quad - \Big(\underbrace{\frac{m_{\max}C_{t-1}}{1 + C_{t-1}/A_{\text{mort,s}}}}_{\text{survival Allee mort.}} + m_{\text{dd}} C_{t-1}^2\Big) + I^{\text{cots}}_t
\end{align}
Observations are lognormal for COTS and normal for coral cover, as implemented.

\paragraph{Parameter table (selected).}
\begin{longtable}{@{}l l l X l l@{}}
\toprule
Symbol & Name & Units & Description & Value & Bounds/Source\\
\midrule
\endhead
$T_{\text{opt,cots}}$ & Opt. temp (COTS repro) & \si{\degreeC} & Gaussian peak temperature & 28.0 & [25,32] (init)\\
$T_{\sigma,\text{cots}}$ & Thermal width & \si{\degreeC} & Gaussian SD & 2.0 & [0.5,5] (init)\\
$r_F,r_S$ & Coral growth rates & \si{yr^{-1}} & Intrinsic growth & \(\exp(\text{log\_r\_fast}), \exp(\text{log\_r\_slow})\) & (lit.)\\
$K_{\text{coral}}$ & Coral capacity & \% & Total coral carrying capacity & \(\exp(\text{log\_K\_coral})\) & (lit.)\\
$a_F,a_S,h$ & Attack/handling & see code & Type II functional response & logs in file & (lit./init)\\
$e_F,e_S$ & Conversion eff. & -- & Coral-to-COTS conversion & logs in file & (lit./init)\\
\bottomrule
\end{longtable}

\clearpage
\section{Best Performing Models for NPZ Case Study}
\label{sec:best_models_npz}

\subsection{openrouter:openai/gpt-5 Model (NPZ)}
\paragraph{States.} Nutrient \(N\), phytoplankton \(P\), zooplankton \(Z\), detritus \(D\) (all \(\si{g\ C\ m^{-3}}\)); time in days.

\paragraph{Limitation and rates.}
\begin{align}
f_T &= q10^{(T_C - T_{\text{ref}})/10},\quad
f_I = \frac{I}{K_I + I},\quad
f_N = \frac{N}{K_N + N} \\
\mu &= \mu_{\max} f_T f_I f_N,\qquad
g = g_{\max} f_T \frac{P^h}{K_G^h + P^h}
\end{align}

\paragraph{Fluxes (per day).}
\begin{align}
U &= \mu P,& G &= g Z,& Z_g &= e_Z G,& R_{g\to D} &= (1-e_Z)G\\
R_{P\to N} &= r_P m_P P,& R_{P\to D} &= (1-r_P)m_P P\\
R_{Z\to N} &= r_Z m_Z Z,& R_{Z\to D} &= (1-r_Z)m_Z Z\\
\mathrm{Ex} &= e_x Z,& \mathrm{Mx} &= k_{\text{mix}}(N_\star - N),\\
\mathrm{Rem} &= k_{\text{rem}} D,& \mathrm{Snk} &= k_{\text{sink}} D
\end{align}

\paragraph{Dynamics (Euler forward).}
\begin{align}
\frac{dN}{dt} &= -U + R_{P\to N} + R_{Z\to N} + \mathrm{Ex} + \mathrm{Mx} + \mathrm{Rem} \\
\frac{dP}{dt} &= U - G - m_P P \\
\frac{dZ}{dt} &= Z_g - m_Z Z - \gamma_Z Z^2 \\
\frac{dD}{dt} &= R_{g\to D} + R_{P\to D} + R_{Z\to D} - \mathrm{Rem} - \mathrm{Snk}
\end{align}

\paragraph{Observation model (lognormal).}
\begin{align}
\log N^{\text{obs}}_i &\sim \mathcal{N}\big(\log N_i,\; \sigma_N^2\big),\quad
\log P^{\text{obs}}_i \sim \mathcal{N}\big(\log P_i,\; \sigma_P^2\big),\quad
\log Z^{\text{obs}}_i \sim \mathcal{N}\big(\log Z_i,\; \sigma_Z^2\big).
\end{align}

\paragraph{Parameters.}
\begin{longtable}{@{}l l l X l l@{}}
\toprule
Symbol & Name & Units & Description & Value & Bounds / Source\\
\midrule
\endhead
$\log\mu_{\max}$ & Max phyto growth (log) & \si{d^{-1}} (log) & At reference conditions & $-0.0204$ & [\(-0.2231, 0.1823\)] (lit.)\\
$\log K_N$ & Nutrient half-sat (log) & \si{g\ C\ m^{-3}} (log) & Uptake half-saturation & $-2.9957$ & [\(-6.9078, 0\)]\\
$I$ & Irradiance proxy & \si{W m^{-2}} & Mean photosynthetic light & 150 & [0, 500] (init)\\
$\log K_I$ & Light half-sat (log) & \si{W m^{-2}} (log) & Light saturation & 4.3175 & [0, 5.7038]\\
$\log g_{\max}$ & Max grazing (log) & \si{d^{-1}} (log) & Per Z biomass & $-0.6931$ & [\(-2.9957, 0.6931\)]\\
$\log K_G$ & Grazing half-sat (log) & \si{g\ C\ m^{-3}} (log) & Holling III half-sat & $-2.3026$ & [\(-6.9078, 0\)]\\
$h$ & Holling exponent & -- & Type III shape & 2.0 & [1, 3] (lit.)\\
$\mathrm{logit}\,e_Z$ & Assimilation (logit) & -- & Zooplankton assimilation & 0 & \(e_Z=0.5\) (lit.)\\
$\log m_P$ & Phyto mortality (log) & \si{d^{-1}} (log) & Linear mortality & $-2.9957$ & [\(-6.9078, -1.2040\)]\\
$\log m_Z$ & Zoo mortality (log) & \si{d^{-1}} (log) & Linear mortality & $-3.5066$ & [\(-6.9078, -1.2040\)]\\
$\log\gamma_Z$ & Zoo self-limit (log) & \((\si{g\ C\ m^{-3}})^{-1}\si{d^{-1}}\) (log) & Quadratic limitation & $-4.6052$ & [\(-9.2103, -1.6094\)]\\
$\mathrm{logit}\,r_P$ & P mort.\,\(\to N\) frac (logit) & -- & Split to nutrient & 0.8473 & (lit.)\\
$\mathrm{logit}\,r_Z$ & Z mort.\,\(\to N\) frac (logit) & -- & Split to nutrient & 0.8473 & (lit.)\\
$\log e_x$ & Excretion (log) & \si{d^{-1}} (log) & Z excretion to N & $-4.6052$ & [\(-13.8155, -1.6094\)]\\
$\log k_{\text{mix}}$ & Mixing rate (log) & \si{d^{-1}} (log) & Nutrient supply & $-3.9120$ & [\(-13.8155, -0.6931\)]\\
$N_\star$ & Deep/source nutrient & \si{g\ C\ m^{-3}} & Mixing target & 0.3 & [0, 2] (init)\\
$\log q10$ & Q10 (log) & -- & Temperature coefficient & 0.659 & [0.606, 0.711] (lit.)\\
$T_C,T_{\text{ref}}$ & Ambient/reference T & \si{\degreeC} & For Q10 scaling & 15, 15 & [0, 35] (init/lit.)\\
$\log k_{\text{rem}}$ & Remineralization (log) & \si{d^{-1}} (log) & Detritus \(\to N\) & $-2.3026$ & [\(-4.6052, 0\)] (added)\\
$\log k_{\text{sink}}$ & Sinking (log) & \si{d^{-1}} (log) & D export & $-4.6052$ & [\(-13.8155, 0\)] (added)\\
$\log\sigma_{N,P,Z}$ & Obs.\ SDs (log) & log-SD & Lognormal errors & $-2.3026$ each & [\(-5, 2\)] (init)\\
\bottomrule
\end{longtable}

\subsection{openrouter:anthropic/claude-sonnet-4.5 Model (NPZ)}
\paragraph{Core processes with variable assimilation efficiency.}
\begin{align}
\text{Seasonal surface light: } I_0(t) &= I_{0,\text{mean}}\Big[1 + A_{I}\sin\!\Big(\frac{2\pi (t - t_{\text{phase}})}{365}\Big)\Big] \\[2pt]
\text{Depth-avg light: } I_{\text{avg}} &= \frac{I_0(t)\,\big(1-e^{-(k_w + k_c P)z_{\text{mix}}}\big)}{(k_w + k_c P)z_{\text{mix}}} \\
\text{Limitations: } \ell_I &= \frac{I_{\text{avg}}}{K_I + I_{\text{avg}}},\quad \ell_N = \frac{N}{K_N + N} \\
\text{Q10 factors: } f_{\text{phy}} &= Q10_{\text{phyto}}^{(T - T_{\text{ref}})/10},\quad
f_{\text{zoo}} = Q10_{\text{zoo}}^{(T - T_{\text{ref}})/10},\quad
f_{\text{rem}} = Q10_{\text{rem}}^{(T - T_{\text{ref}})/10} \\[4pt]
\text{Phyto growth: } G_P &= r_{\max}\, f_{\text{phy}}\, \ell_N\, \ell_I\, P \\
\text{Grazing (Type II): } \mathrm{Graze} &= g_{\max}\, f_{\text{zoo}}\, \frac{P}{K_P + P}\, Z \\
\text{Variable assimilation: } \varepsilon_{\text{eff}} &= \varepsilon_{\min} + (\varepsilon_{\max} - \varepsilon_{\min})\,\ell_N \\[4pt]
\frac{dN}{dt} &= N_{\text{input}} - G_P + \gamma_P f_{\text{rem}}\, m_P P + \gamma_Z f_{\text{rem}}\, m_Z Z + (1-\varepsilon_{\text{eff}})\,\mathrm{Graze} \\
\frac{dP}{dt} &= G_P - m_P f_{\text{rem}}\, P - \mathrm{Graze} \\
\frac{dZ}{dt} &= \varepsilon_{\text{eff}}\, \mathrm{Graze} - \big(m_Z f_{\text{zoo}} Z + m_{Z2} Z^2\big)
\end{align}

\paragraph{Observation model.}
Normal errors on \(N,P,Z\) as specified (see file).

\paragraph{Parameter table (selected).}
\begin{longtable}{@{}l l l X l l@{}}
\toprule
Symbol & Name & Units & Description & Value & Bounds / Source\\
\midrule
\endhead
$r_{\max}$ & Max phyto growth & \si{d^{-1}} & Growth at reference \(T\) & 1.0 & [0.8, 1.2] (lit.)\\
$K_N$ & Nutrient half-sat & \si{g\ C\ m^{-3}} & Uptake saturation & 0.1 & [0, 5] (lit.)\\
$m_P$ & Phyto mortality & \si{d^{-1}} & Non-grazing losses & 0.05 & [0, 1] (lit.)\\
$g_{\max}$ & Max grazing & \si{d^{-1}} & Zoo grazing at ref \(T\) & 0.5 & [0, 5] (lit.)\\
$K_P$ & Grazing half-sat & \si{g\ C\ m^{-3}} & Holling II half-sat & 0.1 & [0, 5] (lit.)\\
\(\varepsilon_{\min},\varepsilon_{\max}\) & Assimilation bounds & -- & Food-quality dependent & 0.15; 0.45 & [0.05–0.40; 0.20–0.70] (lit.)\\
$m_Z,m_{Z2}$ & Zoo mort. (lin/quad) & \si{d^{-1}}, \si{d^{-1}(g\ C\ m^{-3})^{-1}} & Linear + quadratic & 0.05; 0.1 & (lit.)\\
$\gamma_P,\gamma_Z$ & Recycling eff. & -- & Fractions to \(N\) & 0.5; 0.3 & (lit.)\\
$N_{\text{input}}$ & External input & \si{g\ C\ m^{-3}\ d^{-1}} & Upwelling/mixing/depo & 0.01 & (init)\\
$I_{0,\text{mean}},A_{I},t_{\text{phase}}$ & Light params & see units & Seasonal PAR & 443; 0.4; 172 & (lit.)\\
$K_I,k_w,k_c,z_{\text{mix}}$ & Light limitation & see units & Half-sat \& attenuation & 50; 0.04; 0.03; 20 & (lit.)\\
$Q10_{\text{phyto}},Q10_{\text{zoo}},Q10_{\text{rem}}$ & Temperature Q10 & -- & Q10 coefficients & 2.7; 2.5; 3.0 & (lit.)\\
$T_{\text{ref}}$ & Reference temperature & \si{\degreeC} & Q10 reference & 15 & (lit.)\\
\bottomrule
\end{longtable}
