
\section{Comparative Analysis of Best-Performing Models}
\label{sec:model_comparison}

Before presenting the full code for each model, we analyze the key differences between the best-performing models to understand their ecological approaches and mathematical structures.

% \input{supp_comparison}

\begin{landscape}
    \footnotesize
    \setlength{\tabcolsep}{3pt}
    \renewcommand{\arraystretch}{1.2}
    
    \begin{longtable}{@{}P{0.8\fivecol}P{\fivecol}P{\fivecol}P{\fivecol}P{\fivecol}@{}}
    \caption{Detailed Ecological Mechanisms\label{tab:ecological_mechanisms}}\\
    \toprule
    \textbf{Mechanism} & \textbf{Human} & \textbf{Gemini 2.5 Pro} & \textbf{GPT-5} & \textbf{Sonnet 4.5} \\
    \midrule
    \endfirsthead
    \multicolumn{5}{c}{\bfseries Detailed Ecological Mechanisms — continued}\\
    \toprule
    \textbf{Mechanism} & \textbf{Human} & \textbf{Gemini 2.5 Pro} & \textbf{GPT-5} & \textbf{Sonnet 4.5} \\
    \midrule
    \endhead
    \midrule
    \multicolumn{5}{r}{Continued on next page}\\
    \endfoot
    \bottomrule
    \endlastfoot
    
    \textbf{COTS state structure / life history} &
\textit{Age-structured (3 classes)}: \(N_{t+1,1} = N_{t,0} e^{-M_0}\); \(N_{t+1,2} = N_{t,1} e^{-f M_1} + N_{t,2} e^{-f M_2}\); recruits form \(N_{t+1,0}\). Age-specific mortality \(M_a = M_{\text{cots}} + \lambda /(1+a)\). &
\textit{Single compartment} \(C_t\) (no explicit ages). Growth from consumption + modifiers. &
\textit{Two-stage} (juveniles \(J\) and adults \(C\)): maturation \(m_J J\); juvenile mortality \(\mu_J J\); adult mortality \((\mu_C + \gamma_C C)C\). &
\textit{Single compartment} with logistic-like adult growth plus additive recruitment pulse; no age classes. \\
\midrule
\textbf{COTS stock--recruitment} &
\textit{Beverton--Holt (BH) from spawners}:
\(R_{t+1} = \dfrac{\alpha (N_{t+1,2}/K_{\text{sp}})}{\beta + (N_{t+1,2}/K_{\text{sp}})}\),
with \(\alpha,\beta\) derived from slope steepness \(h\) and \(R_0\). &
\textit{Not BH}:
recruitment implicit via \(\big(e_F \cdot \text{Cons}_F + e_S \cdot \text{Cons}_S\big)\)
\(\times\) Allee \(\times\) temperature Gaussian (no explicit SR function). &
\textit{BH-like taper on adults}:
\(\text{Rec} = \alpha_{\text{rec}}\,[ C^\phi /(1 + C/C_{\text{sat}} ) ] \cdot f_{\text{Allee}}(C) \cdot f_T(\text{SST}) \cdot f_{\text{food}}
+ \text{imm}_{t-1}\). &
\textit{Additive pulse recruitment}:
\(\text{recruit\_pulse} = R_{\max}\cdot \sigma\!\big(\text{fav}-\tau\big)\cdot \text{fav}\),
added to adult change; adult growth also logistic-like. \\
\midrule
\textbf{COTS immigration \& interannual pulses} &
\textit{Background immigration} + year-specific deviations:
\(N_{t+1,0} = (R_{t+1} + \text{Imm}_{\text{CoTS}})\exp(\varepsilon_{t+1} + \sigma^2/2)\);
\(\varepsilon_t\) applied in specified years. &
Exogenous additive series \(\text{cotsimm}(t)\) added directly to \(C_t\) dynamics. &
Exogenous series \(\text{cotsimm}_{t-1}\) added to recruitment (lagged). &
Used twice:
(i) normalized within favorability index, and
(ii) multiplicative \(\text{immigration\_boost} = 1 + \text{effect}\cdot \text{imm}\) on adult growth. \\
\midrule
\textbf{COTS mortality and its drivers} &
\textit{Baseline + age-dependent}, modulated by coral availability:
effective multiplier \(f = (1 - \tilde p) + \tilde p \cdot \rho\) (see prey switching \(\rho\));
applied to age-1 and age-2+ mortality; mortality reduced when fast coral abundant. &
\textit{No age structure};
mortality = survival Allee
\(\dfrac{m_{C,\max} C}{1 + C/A_{\text{mort}}}\)
\(+\) quadratic density dependence \(m_{C,\text{dd}} C^2\);
not explicitly coral-modulated. &
\textit{No age structure};
adult mortality \((\mu_C + \gamma_C C) C\);
not explicitly coral-modulated. &
\textit{No age structure};
baseline + density-dependent mortality,
amplified by starvation:
multiplier \(1 + 2\,e^{- (F+S)/5}\) increases mortality when coral scarce. \\
\midrule
\textbf{Prey switching / diet preference} &
\textit{Abundance-driven switching}:
\(\rho = \exp(-\text{switchSlope}\cdot F/K)\).
Predation share on fast coral \(\propto (1-\rho)\), on slow \(\propto \rho\).
As fast coral increases, switch toward fast prey. &
Implicit via multi-prey Type II with separate \(a_F, a_S\); no explicit \(\rho\) rule. &
Implicit via Type II/III blend (exponents \(\eta_F,\eta_S\) create low-prey refuge); no explicit \(\rho\). &
\textit{Preference + availability}:
weight combines fixed preference with fast-coral proportion (soft switching toward abundant prey). \\
\midrule
\textbf{Functional response (COTS $\rightarrow$ coral consumption)} &
\textit{Sigmoid-saturated vs COTS density}:
\(Q_F = F\,(1-\rho)\,p_{1F}\,\dfrac{N_{1+2}}{1+\exp\!\big(-N_{1+2}/p_{2F}\big)}\),
\(Q_M = S\,\rho\,p_{1M}\,\dfrac{N_{1+2}}{1+\exp\!\big(-N_{1+2}/p_{2M}\big)}\). &
\textit{Multi-prey Holling Type II}:
per-capita
\(\dfrac{a_F F}{1 + a_F h F + a_S h S}\),
\(\dfrac{a_S S}{1 + a_F h F + a_S h S}\);
totals scale with \(C\). &
\textit{Type II/III blend}:
\(\text{Cons}_F = q_F \dfrac{a_F F^{\eta_F} C}{1 + h(a_F F^{\eta_F} + a_S S^{\eta_S})}\),
\(\text{Cons}_S = q_S \dfrac{a_S S^{\eta_S} C}{1 + h(a_F F^{\eta_F} + a_S S^{\eta_S})}\). &
\textit{Type II per prey} with separate handling;
then \textit{preference weighting} mixes fast vs slow consumption. \\
\midrule
\textbf{Coral intrinsic growth \& space competition} &
\textit{Logistic regrowth with shared space}:
\(\displaystyle F_{t+1} = F_t \Big[1 + \rho_F(\text{SST})\, r_f \Big(1 - \tfrac{F_t + S_t}{K}\Big)\Big] - Q_F - M_{\text{ble},F}\);
analogous for \(S\). &
\textit{Logistic regrowth} with shared \(K_{\text{coral}}\);
losses: predation + bleaching mortality. &
\textit{Logistic regrowth} with shared \(K_{\text{tot}}\);
losses: predation + heat-related terms. &
\textit{Logistic regrowth} with shared \(K_{\text{coral}}\);
losses: predation + temperature-stress mortality. \\
\midrule
\textbf{SST modulation of coral regrowth (performance curve)} &
\textit{Gaussian performance multiplier}:
\(\rho_F(\text{SST}) = \exp\!\left[-\dfrac{(\text{SST} - \text{SST0}_f)^2}{2\,\text{SST\_sig}_f^2}\right]\),
\(\rho_M(\text{SST}) = \exp\!\left[-\dfrac{(\text{SST} - \text{SST0}_m)^2}{2\,\text{SST\_sig}_m^2}\right]\). &
Not Gaussian on growth; SST enters via \textit{logistic bleaching mortality} (see below). &
\textit{Heat-stress growth reduction}:
multiplier \(\exp[-\beta_{\text{bleach}} \max(0, \text{SST} - T_{\text{bleach}})]\) on growth (non-Gaussian). &
\textit{Threshold stress loss}:
linear mortality above \(T_{\text{stress}}\); no Gaussian growth multiplier. \\
\midrule
\textbf{Coral bleaching mortality} &
\textit{Logistic bleaching}:
\(M_{\text{ble},F} = F \cdot \big[1 + \exp\{-\eta_f(\text{SST} - M_{50,F})\}\big]^{-1}\);
analogous for slow coral (impulse option commented). &
\textit{Logistic bleaching mortality}:
\(m_{F,\text{sst}} / \big[1 + \exp\{-k_{\text{bleach}}(\text{SST} - T_{\text{bleach},F})\}\big]\);
applied proportionally to \(F\); analogous for \(S\). &
Two components:
(i) multiplicative \textit{growth reduction} under heat,
(ii) \textit{additional linear loss} \(m_{\text{bleach}} \cdot \text{heat} \cdot \text{coral}\);
no logistic 50\% curve. &
\textit{Temperature-stress mortality} proportional to degrees above threshold;
no explicit logistic bleaching curve. \\
\midrule
\textbf{Coral carrying capacity (space sharing)} &
\textit{Shared \(K\)} via \(1 - (F+S)/K\) in both coral equations. &
\textit{Shared \(K_{\text{coral}}\)} in coral logistic growth. &
\textit{Shared \(K_{\text{tot}}\)} in coral logistic growth. &
\textit{Shared \(K_{\text{coral}}\)} in coral logistic growth. \\
\bottomrule

% \caption{Side-by-side comparison of ecological mechanisms and mathematical implementations across the human-generated reference model (\texttt{CoTSmodel\_v4.cpp}) and three LLM-generated models. The rows are anchored to mechanisms explicitly present in the human model; each LLM implementation is summarized without parameter values. Symbols: \(C\) adult COTS, \(J\) juvenile COTS, \(F,S\) fast/slow coral cover, \(K\) shared coral carrying capacity, \(\text{SST}\) sea surface temperature.}
\label{tab:mechanism_comparison}
\end{longtable}
\end{landscape}

\begin{landscape}
    \footnotesize
    \renewcommand{\arraystretch}{1.15}
    \setlength{\tabcolsep}{3pt}
    
    % Equal-width columns: first column slightly wider
    \begin{longtable}{@{}P{0.8\fivecol}P{\fivecol}P{\fivecol}P{\fivecol}P{\fivecol}@{}}
    \caption{Model Parameterisation Comparison\label{tab:human_roles_models}}\\
    \toprule
    \textbf{Human (name)} & \textbf{Role / Units} & \textbf{Gemini 2.5 Pro} & \textbf{OpenAI GPT--5} & \textbf{Claude Sonnet 4.5} \\
    \midrule
    \endfirsthead
    
    \multicolumn{5}{c}{\bfseries Model Parameterisation Comparison — continued}\\
    \toprule
    \textbf{Human (name)} & \textbf{Role / Units} & \textbf{Gemini 2.5 Pro} & \textbf{OpenAI GPT--5} & \textbf{Claude Sonnet 4.5} \\
    \midrule
    \endhead
    
    \midrule
    \multicolumn{5}{r}{Continued on next page}\\
    \endfoot
    
    \bottomrule
    \endlastfoot
    
    % --- Rows ---
\texttt{Mcots} & Baseline instantaneous COTS mortality (yr$^{-1}$) &
$m_{C,\max}$ (low-density mortality scale); $m_{C,\text{dd}}$ (DD mortality) &
$\mu_C$ (baseline adult mortality), $\gamma_C$ (DD mortality) &
$\texttt{log\_mort\_base}$ (baseline), $\texttt{log\_mort\_density}$ (DD) \\
\texttt{lam} & Age-dependence of mortality, $M_a = M_{\rm cots}+\lambda/(1+a)$ &
--- (no age structure) & --- (two-stage but no age-specific $M$) & --- (no age structure) \\
\texttt{ptil} & Fraction of $M$ attributable to fast-coral availability (mortality modulation) &
--- (no coral-modulated $M$) &
--- (no coral-modulated $M$) &
Starvation multiplier $1+2e^{-(F+S)/5}$ (different form) \\
\texttt{h} (BH steepness) & Shapes SR via $(R_0,h)\rightarrow \alpha,\beta$ &
--- (no BH SR; growth from consumption) &
$C_{\text{sat,rec}}$ (BH-like taper), $\alpha_{\rm rec}$, $\phi$ (fecundity exponent) &
Recruitment pulse parameters: $\texttt{log\_recruit\_max}$, $\texttt{recruit\_threshold}$ \\
\texttt{R0} & Recruitment at unexploited state (for SR derivation) &
--- (no explicit $R_0$) &
$\alpha_{\rm rec}$ (scales juvenile input; closest analogue) &
$\texttt{log\_recruit\_max}$ (caps pulse magnitude; different structure) \\
\texttt{Imm\_CoTS} & Background immigration (ind m$^{-2}$ yr$^{-1}$) &
$\texttt{cotsimm\_dat}(t)$ (added to adults each step) &
$\texttt{cotsimm\_dat}(t\!-\!1)$ (added into recruitment) &
$\texttt{cotsimm\_dat}$ used in favorability and as growth boost \\
\texttt{sigCoTS} & SR process variability (lognormal on recruits) &
--- (no process noise term; observation SD only) &
--- (no explicit process noise on SR; observation SDs) &
--- (no explicit process noise on SR; observation SDs) \\
\texttt{immigration} & Year-specific recruitment deviations (vector by year) &
Represent via time series in $\texttt{cotsimm\_dat}$ &
Represent via time series in $\texttt{cotsimm\_dat}$ &
Represent via time series in $\texttt{cotsimm\_dat}$ \\
\texttt{COTS\_init} & Initial COTS abundance (age-2+) &
Init from first data row (no param) &
$C0$ (adults), $J0$ (juveniles) &
Init from first data row (no param) \\
\texttt{p1f}, \texttt{p1m} & Per-COTS coral loss coefficients (fast/slow), units of \% cover $\cdot$ (ind$^{-1}$ yr$^{-1}$) &
$a_F,a_S$ (attack), $h$ (handling) with efficiencies $e_F,e_S$ for COTS growth (coral loss $\propto$ consumption) &
$aF,aS$ (attack), $h$ (handling), $qF,qS$ (loss efficiencies) \quad(Type II/III via $\eta_F,\eta_S$) &
$\texttt{log\_attack\_fast/slow}$, $\texttt{log\_handling\_fast/slow}$, plus preference weighting \\
\texttt{p2f}, \texttt{p2m} & Logistic saturation vs COTS density (shape of $Q_F,Q_M$) &
Denominator $1+a_FhF+a_ShS$ (resource-based saturation; no explicit p2) &
Denominator $1+h(a_FF^{\eta_F}+a_SS^{\eta_S})$ (no explicit p2) &
Separate Type-II denominators per prey (no explicit p2) \\
\texttt{switchSlope} & Controls prey switching: $\rho=\exp(-\text{switchSlope}\cdot F/K)$ &
--- (no explicit $\rho$) &
$\eta_F,\eta_S$ (Type-III curvature; soft switching), no $\rho$ &
$\texttt{preference\_fast}$ blended with fast-coral share (soft switching) \\
\texttt{K} & Shared coral carrying capacity (\% cover scale) &
$K_{\text{coral}}$ (\texttt{log\_K\_coral} parameterized) &
$K_{\text{tot}}$ &
$K_{\text{coral}}$ (\texttt{log\_K\_coral}) \\
\texttt{rf}, \texttt{rm} & Intrinsic regrowth (fast/slow), yr$^{-1}$ &
$r_F,r_S$ (\texttt{log\_r\_F}, \texttt{log\_r\_S}) &
$rF,rS$ &
$r_{\text{fast}},r_{\text{slow}}$ (\texttt{log\_r\_fast}, \texttt{log\_r\_slow}) \\
\texttt{Cf\_init}, \texttt{Cm\_init} & Initial coral state (fraction of $K$) &
Init from first data row (no param) &
$F0,S0$ &
Init from first data row (no param) \\
\texttt{Eta\_f}, \texttt{Eta\_m} & Logistic bleaching slope (fast/slow) &
$k_{\text{bleach}}$ (common steepness) &
$\beta_{\text{bleach}}$ (growth reduction) and $m_{\text{bleachF/S}}$ (linear loss) &
$\texttt{temp\_stress\_rate}$ (linear loss; no logistic) \\
\texttt{M\_SST50\_f}, \texttt{M\_SST50\_m} & 50\% bleaching SST (fast/slow) &
$T_{\text{bleach},F/S}$ &
$T_{\text{opt,bleach}}$ (single threshold) &
$\texttt{temp\_stress\_threshold}$ (single threshold) \\
\texttt{Ble\_imp\_f}, \texttt{Ble\_imp\_m} & Optional impulse bleaching toggles &
--- &
--- &
--- \\
\texttt{SST0\_f}, \texttt{SST0\_m} & Coral Gaussian performance optima (fast/slow) &
--- (no Gaussian growth multiplier; bleaching only) &
$T_{\text{opt,bleach}}$ used in heat reduction (not Gaussian peak) &
--- (no Gaussian growth multiplier) \\
\texttt{SST\_sig\_f}, \texttt{SST\_sig\_m} & Coral Gaussian performance widths (fast/slow) &
--- & --- & --- \\
\addlinespace
--- &
--- &
$T_{\text{opt,rec}}, \beta_{\text{rec}}$ (Gaussian on recruitment) &
$\texttt{log\_temp\_opt},\ \texttt{log\_temp\_width}$ (Gaussian in favorability \& adult growth) &
$T_{\text{opt,cots}}, T_{\sigma,\text{cots}}$ (Gaussian on COTS reproduction) \\
\label{tab:params}
\end{longtable}
\end{landscape}
